Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields

### 2017-07-26 Master Thesis/synopsis.tex
Add Lemma before 2.4 and remove reference to [K1] 1.4

@@

Let $\mathfrak{X}$ be a model of $K$. For $x\in\mathfrak{X}_1$ the local ring $\mathcal{O}_{\mathfrak{X}, x}$ is a discrete valuation ring and defines a valuation $v_x\in\operatorname{Spv}_\text{d}^{(1)}(K)$. By \cite{Ka86} Theorem 0.8, the map
\[ \mathrm{H}^3(K)\longrightarrow \prod_{x\in\mathfrak{X}_1}\mathrm{H}^3(K_{v_x}) \]
is injective, and, in fact, the proof shows that its image is contained in $\bigoplus \mathrm{H}^3(K_{v_x})$. By \cite{Ka86} Lemma 1.4, for a discrete valuation $v$ of $K$ the map
\[ \mathrm{H}^3(K_v) \longrightarrow \prod_w\mathrm{H}^3(K_w) \]
where $w$ ranges over all rank-$2$ valuations with $v(w) = v$, can be identified with the following map:
\[ {}_\ell\operatorname{Br}(k) \longrightarrow \prod_{\widetilde{w}\in\operatorname{Spv}^{(1)}(k)} {}_\ell\operatorname{Br}(k_{\widetilde{w}}) \]
where $k = k(v)$. This map is injective by the classical Hasse principle and its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\widetilde{w}})$. Combining these maps yields the injection:
\[\Phi: \mathrm{H}^3(K)\longrightarrow \bigoplus_{x\in\mathfrak{X}_1}\  \bigoplus_{\substack{w\in\operatorname{Spv}_\text{rd}^{(2)}(K) \\ v(w) = v_x}} \mathrm{H}^3(K_w) \]
The direct sum ranges over a subset of $|\operatorname{Spv}(K)|$, thus this concludes the proof. 
\end{proof}

@@

Let $G_K(\ell)$ be the maximal pro-$\ell$ quotient of $G_K$. Since $\mu_\ell\subset K$, the action of $G_k$ on the $\ell$-Sylow group of the inertia group of $w$ factors through the maximal pro-$\ell$ quotient $G_k(\ell)$ of $G_k$. Since $G_{k,\ell}\cong G_k(\ell)$, we also see that the canonical projection $G_K\to G_K(\ell)$ restricts to an isomorphism $G_\ell\to G_K(\ell)$. Now consider the following composition:
\[ \begin{tikzcd}
\mathrm{H}^3(G_K(\ell),\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{inf}"] & \mathrm{H}^3(G_K,\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{res}"] & \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})
\end{tikzcd} \]
Since $\operatorname{res}\circ\operatorname{inf}$ is an isomorphism, $\operatorname{inf}$ is injective and $\operatorname{res}$ is surjective. We have $\operatorname{cor}\circ\operatorname{res} = (G_K:G_\ell)$. However, $(G_K:G_\ell)$ is coprime to $\ell$, which makes $\operatorname{res}$ injective as well. Therefore, $\operatorname{inf}$ and $\operatorname{res}$ are both isomorphisms and it suffices to calculate $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})$. By \cite{Se64} Ch. I, §4 Proposition 22 we have
\[\operatorname{cd}(G_\ell) = \operatorname{cd}(\mathbb{Z}_\ell^r\rtimes G_k(\ell)) = \begin{cases}
r, & G_k(\ell) = 0 \\
r + 1, & G_k(\ell)\cong\mathbb{Z}_\ell
\end{cases} \] 
since $\operatorname{cd}(\mathbb{Z}_\ell)=1$. Thus $\mathrm{H}^3(G_\ell, \mathbb{Z}/\ell\mathbb{Z}) =0$ iff $G_k(\ell) = 0$ or $r < 2$. Now let $G_k(\ell) \cong\mathbb{Z}_\ell$ and $r = 2$. According to the proof of \cite{Se64} Ch. I, §4 Proposition 22, we have:
\[ \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) = \mathrm{H}^1(G_k(\ell), \mathrm{H}^2 (\mathbb{Z}_\ell^r, \mathbb{Z}/\ell\mathbb{Z})) \]
By [NSW] Theorem 2.4.6, we have $\mathrm{H}^2(\mathbb{Z}_\ell^r,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^1(\mathbb{Z}_\ell, \mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})) = \mathbb{Z}/\ell\mathbb{Z}$, since $\mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\operatorname{Hom}_\text{cont.}(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$. Therefore, $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$.

The second assertion follows from the first one, applied to $L$, by inspecting the following equation of supernatural numbers (\cite{En72} Theorem 20.21):
\[ [L:K] = (\Gamma':\Gamma) \cdot [k':k]\cdot p^d \]
for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique prolongation of $w$ to $L$. By using this equation, it suffices to show the following equivalences:
\begin{enumerate}[(i)]
\item $\ell^\infty \nmid [k':k] \iff \ell^\infty \nmid [k':\mathbb{F}_p]$
\item $\ell^\infty\nmid (\Gamma':\Gamma) \iff \dim_{\mathbb{F}_\ell}(\Gamma'/\ell\Gamma')=2$
\end{enumerate}
(i) is clear. For (ii) let $T'$ be the absolute inertia group of $L$. By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T'$ is isomorphic to $\mathbb{Z}_\ell^{r'}$, where $r'=\dim_{\mathbb{F}_\ell}(\Gamma'/\ell\Gamma')$. Furthermore, we have $(\Gamma':\Gamma) = (T:T')$ and the following exact sequence of $\ell$-Sylow groups:
\[ 0 \longrightarrow \mathbb{Z}_\ell^{r'} \longrightarrow \mathbb{Z}_\ell^{r}\longrightarrow (T/T')_\ell\longrightarrow 0 \]
Now $(T/T')_\ell$ is finite if and only if $r'=r=2$.
\end{proof}

>>>>

\begin{lemma}\label{2.4-pre}
Let $K$ be henselian with respect to a discrete valuation $v$, $k=k(v)$ its residue field with $\operatorname{cd}(k)\leq 2$. Let $\ell\neq p$ be a prime where $p=\operatorname{char}(k)$. Then:
\[ \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}(2)) \cong \mathrm{H}^2(k, \mu_{\ell}) \cong {}_\ell\operatorname{Br}(k) \]
\end{lemma}

\begin{proof}
Let $T$ be the absolute inertia group of $K$ and $V\lhd T$ its absolute ramification group. By Theorem \ref{theorem-sylowgroups-inertiagroup}, $V$ is the $p$-Sylow group of $T$ and the $\ell$-Sylow group of $T/V$ is isomorphic to $\mathbb{Z}_\ell$. Therefore, by \cite{Se64} Ch. I, §3 Proposition 15:
\[ \operatorname{cd}_\ell(T)\leq \operatorname{cd}_\ell(V) + \operatorname{cd}_\ell(\mathbb{Z}_\ell) = 1 \]
Inspection of the Hochschild-Serre spectral sequence 
\[E_2^{i,j}=\mathrm{H}^i(G_K/T, \mathrm{H}^j(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\Longrightarrow\mathrm{H}^{i+j}(G_K, \mathbb{Z}/\ell\mathbb{Z}(2))\]
yields an isomorphism $\mathrm{H}^2(G_K/T, \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\cong \mathrm{H}^3(G_K, \mathbb{Z}/\ell\mathbb{Z}(2))$. By Theorem \ref{theorem-absoluteinertia}, we have  $G_K/T\cong G_k$. Since $\mu_{\ell}$ is contained in the absolute inertia field of $K$, $T$ acts trivially on $\mu_{\ell}$, hence:
\[ \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)) \cong \mathrm{H}^1(T, \mu_{\ell}) \cong \operatorname{Hom}_\text{cont.}(T, \mu_{\ell}) \cong \operatorname{Hom}_\text{cont.}(T/V, \mu_{\ell}) \]
By Theorem \ref{theorem-sylowgroups-inertiagroup}, $T/V$ is the free pro-$\mathcal{C}$ group, where $\mathcal{C}$ is the class of all finite groups with order prime to $p$. Therefore $\operatorname{Hom}_\text{cont.}(T/V, \mu_{\ell})\cong \mu_{\ell}$. The rest follows from Proposition \ref{prop:kummer-theory}.
\end{proof}

@@

Let $\mathfrak{X}$ be a model of $K$. For $x\in\mathfrak{X}_1$ the local ring $\mathcal{O}_{\mathfrak{X}, x}$ is a discrete valuation ring and defines a valuation $v_x\in\operatorname{Spv}_\text{d}^{(1)}(K)$. By \cite{Ka86} Theorem 0.8, the map
\[ \mathrm{H}^3(K)\longrightarrow \prod_{x\in\mathfrak{X}_1}\mathrm{H}^3(K_{v_x}) \]
is injective, and, in fact, the proof shows that its image is contained in $\bigoplus \mathrm{H}^3(K_{v_x})$.

Let $v$ be a discrete valuation of $K$ with residue field $k=k(v)$. By Lemma \ref{2.4-pre}, the map
\[ \mathrm{H}^3(K_v) \longrightarrow \prod_w\mathrm{H}^3(K_w) \]
where $w$ ranges over all rank-$2$ valuations with $v(w) = v$, can be identified with the following map:
\[ {}_\ell\operatorname{Br}(k) \longrightarrow \prod_{\widetilde{w}\in\operatorname{Spv}^{(1)}(k)} {}_\ell\operatorname{Br}(k_{\widetilde{w}}) \]
This map is injective by the classical Hasse principle and its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\widetilde{w}})$. Combining these maps yields the injection:
\[\Phi: \mathrm{H}^3(K)\longrightarrow \bigoplus_{x\in\mathfrak{X}_1}\  \bigoplus_{\substack{w\in\operatorname{Spv}_\text{rd}^{(2)}(K) \\ v(w) = v_x}} \mathrm{H}^3(K_w) \]
The direct sum ranges over a subset of $|\operatorname{Spv}(K)|$, thus this concludes the proof. 
\end{proof}

@@

Let $G_K(\ell)$ be the maximal pro-$\ell$ quotient of $G_K$. Since $\mu_\ell\subset K$, the action of $G_k$ on the $\ell$-Sylow group of the inertia group of $w$ factors through the maximal pro-$\ell$ quotient $G_k(\ell)$ of $G_k$. Since $G_{k,\ell}\cong G_k(\ell)$, we also see that the canonical projection $G_K\to G_K(\ell)$ restricts to an isomorphism $G_\ell\to G_K(\ell)$. Now consider the following composition:
\[ \begin{tikzcd}
\mathrm{H}^3(G_K(\ell),\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{inf}"] & \mathrm{H}^3(G_K,\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{res}"] & \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})
\end{tikzcd} \]
Since $\operatorname{res}\circ\operatorname{inf}$ is an isomorphism, $\operatorname{inf}$ is injective and $\operatorname{res}$ is surjective. We have $\operatorname{cor}\circ\operatorname{res} = (G_K:G_\ell)$. However, $(G_K:G_\ell)$ is coprime to $\ell$, which makes $\operatorname{res}$ injective as well. Therefore, $\operatorname{inf}$ and $\operatorname{res}$ are both isomorphisms and it suffices to calculate $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})$. By \cite{Se64} Ch. I, §4 Proposition 22 we have
\[\operatorname{cd}(G_\ell) = \operatorname{cd}(\mathbb{Z}_\ell^r\rtimes G_k(\ell)) = \begin{cases}
r, & G_k(\ell) = 0 \\
r + 1, & G_k(\ell)\cong\mathbb{Z}_\ell
\end{cases} \] 
since $\operatorname{cd}(\mathbb{Z}_\ell)=1$. Thus $\mathrm{H}^3(G_\ell, \mathbb{Z}/\ell\mathbb{Z}) =0$ iff $G_k(\ell) = 0$ or $r < 2$. Now let $G_k(\ell) \cong\mathbb{Z}_\ell$ and $r = 2$. Inspecting the Hochschild-Serre spectral sequence yields an isomorphism:
\[ \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) \cong \mathrm{H}^1(G_k(\ell), \mathrm{H}^2 (\mathbb{Z}_\ell^r, \mathbb{Z}/\ell\mathbb{Z})) \]
Using $\mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})\cong\operatorname{Hom}_\text{cont.}(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$ and \cite{NSW} Theorem 2.4.6, we can calculate $\mathrm{H}^2(\mathbb{Z}_\ell^r,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathbb{Z}/\ell\mathbb{Z}$. Therefore, $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})\cong\mathbb{Z}/\ell\mathbb{Z}$.

For the second assertion, we assume $\ell^\infty\nmid [k:\mathbb{F}_p]$ and $\dim_{\mathbb{F}_\ell}(\Gamma / \ell\Gamma) = 2$. By \cite{En72} Theorem 20.21, we have the following equation of supernatural numbers:
\[ [L:K] = (\Gamma':\Gamma) \cdot [k':k]\cdot p^d \]
for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique extension of $w$ to $L$. By using this equation and the already proven statement from the first part, it suffices to show the following equivalences:
\begin{enumerate}[(i)]
\item $\ell^\infty \nmid [k':k] \iff \ell^\infty \nmid [k':\mathbb{F}_p]$
\item $\ell^\infty\nmid (\Gamma':\Gamma) \iff \dim_{\mathbb{F}_\ell}(\Gamma'/\ell\Gamma')=2$
\end{enumerate}
(i) follows from $[k':\mathbb{F}_p]=[k':k][k:\mathbb{F}_p]$. For (ii) let $T'$ be the absolute inertia group of $L$. By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T'$ is isomorphic to $\mathbb{Z}_\ell^{r'}$, where $r'=\dim_{\mathbb{F}_\ell}(\Gamma'/\ell\Gamma')$. Furthermore, we have $(\Gamma':\Gamma) = (T:T')$ and the following exact sequence of $\ell$-Sylow groups:
\[ 0 \longrightarrow \mathbb{Z}_\ell^{r'} \longrightarrow \mathbb{Z}_\ell^{r}\longrightarrow (T/T')_\ell\longrightarrow 1 \]
Now $(T/T')_\ell$ is finite if and only if $r'=r=2$.
\end{proof}
